#Introduction to Categorical Data Analysis

##Chapter 2: Contingency Tables

###Table 2.1
```{r table 2.1}
library(htmlTable)
x <- matrix(c(509, 116, 398, 104), nrow = 2, ncol = 2, byrow = TRUE)
dimnames(x) <- list("Gender" = c("Females", "Males"), "Belief in Afterlife" = c("Yes", "No or Undecided"))
htmlTable(x)
```

###Table 2.3 & Analysis
```{r table 2.3}
x <- matrix(c(189, 10845, 104, 10933), nrow = 2, ncol = 2, byrow = TRUE)
dimnames(x) <- list("Group" = c("Placebo", "Aspirin"), "Myocardial Infarction" = c("Yes", "No"))
htmlTable(x)
prop.test(x, correct = FALSE)
```
As in the book, we can see that those in the aspirin group have a lesser prob.
of a heart attack ($p_1-p_2>0$ which implies that the risk is higher for placebo)

###Table 2.3 Risk and Odds Ratio Analysis
```{r table 2.3 risk ratio}
library(Epi)
twoby2(x)
```

###Table 2.4 & Analysis
```{r table 2.4}
x <- matrix(c(172, 173, 90, 346), nrow = 2, ncol = 2, byrow = TRUE)
dimnames(x) <- list("Ever Smoker" = c("Yes", "No"), "MI" = c("MI Cases", "Controls"))
htmlTable(x)
prop.table(x, 2) #given MI, prop that smoked
twoby2(x)
```
Thus, we see that the odds of ever having been a smoker are four times higher
for those who have had an MI vs those who have not.

###Table 2.5: Chi-sq and Odds Ratio
```{r table 2.5}
x <- matrix(c(762, 327, 468, 484, 239, 477), nrow = 2, ncol = 3, byrow = TRUE)
dimnames(x) <- list("Gender" = c("Females", "Males"), "Pol Party" = c("Dem", "Ind", "Rep"))
htmlTable(x)
test <- chisq.test(x, correct = FALSE)
test
htmlTable(round(test$stdres, 2))
```
From the results of the chi squared test, we see that political affiliation and
gender are dependent. Furthermore, based on the standardized residuals, we see
that males are disproportionatly more republican and females more democratic than
what we would expect if the variables were truly independent.

```{r table 2.5 odds ratio}
x_dr <- x[,c(1,3)]
twoby2(x_dr)
```
From the odds ratio above, females have 1.6x the odds of identifying democratic, compared
to their male counterpart.

###Table 2.5 Partitioning
```{r table 2.5 parition}
x1 <- x[,c(1,2)]
htmlTable(x1)
x1_test <- chisq.test(x1, correct = FALSE)
x1_test
x2 <- cbind(rowSums(x[,1:2]), x[,3])
dimnames(x2) <- list("Gender" = c("Female", "Male"), "Pol Party" = c("Dem/Ind", "Rep"))
htmlTable(x2)
x2_test <- chisq.test(x2, correct = FALSE)
x2_test
x1_test$statistic + x2_test$statistic
test$statistic
```
From partitioning above, we can see that when comparing democratic and independent,
there is not evidence of dependence, but when these combined are compared to
republican, we see a statistical difference. Finally, when adding the chi-sq
statistics from the two partitioned tests, they are approximately equal to the
overall test. Do note that if a $G^2$ test had been done (ie with `DescTools::GTest`),
then the statistics would exactly sum to the omnibus test.

###Table 2.7 Trend Test
```{r table 2.7 trend}
library(DescTools)
x <- matrix(c(17066, 48, 14464, 38, 788, 5, 126, 1, 37, 1), nrow = 5, ncol = 2, byrow = TRUE)
dimnames(x) <- list("Alcohol Consumption" = c("0", "<1", "1-2", "3-5", ">=6"),
                    "Malformation" = c("Absent", "Present"))
htmlTable(x)
chisq.test(x, correct = FALSE)
GTest(x, correct = "none")
malformations <- c(48, 38, 5, 1, 1)
totals <- c(17114, 14502, 793, 127, 38)
prop.trend.test(x = malformations, n = totals, score = c(0, 0.5, 1.5, 4.0, 7))
#alternatively: 
prop.trend.test(x = x[,2], n = rowSums(x), score = c(0, 0.5, 1.5, 4.0, 7))
```
From the trend test above, we see that although the chi-sq $\chi^2$ and $G^2$ tests gave
contradicting results (due to small sample size in some cells), we could make
use of the ordinality of the alcohol consumption variable to see that as the
number of drinks increase, so does the trend of having a birth defect.

###Table 2.8 Fisher's Exact Test
```{r table 2.8 fisher}
x <- matrix(c(3, 1, 1, 3), nrow = 2, ncol = 2, byrow = TRUE)
dimnames(x) <- list("Poured First" = c("Milk", "Tea"),
                    "Guess Poured First" = c("Milk", "Tea"))
htmlTable(x)
fisher.test(x = x, alternative = "greater")
```
Unfortunately, it does not appear that his colleague was able to discern
whether tea or milk was poured first. However, it is difficult to detect a
difference with such a small sample size, even with tests for small sample
sizes, such as the Fisher's Exact Test.

###Table 2.10 and Conditional Odds Ratios
```{r table 2.10}
x <- array(c(53, 11, 414, 37, 0, 4, 16, 139), dim = c(2, 2, 2),
           dimnames = list("Defendant's Race" = c("White", "Black"),
                           "Death Penalty" = c("Yes", "No"),
                           "Victim's Race" = c("White", "Black")))
x
apply(x, 3, twoby2)[[c(1, 2)]]
```
Above are the two conditional odds ratio output, controlling for the victim's race.
We can see that, given the victim's race is white, the odds of a white defendant receiving the death
penalty is 43% that of a black defendant. Furthermore, when the victim's race is
black, we see that the odds of a white defendant being handed the death penalty
is 0, because the death penalty was never given to a white defendant when the victim
was black. Something of note is that the marginal odds ratio (collapsing over victim race)
is 1.45 for whites, which implies that whites have higher odds of getting the
death penalty (Simpson's Paradox).

###Table 2.11 and Conditional Independence
```{r table 2.11}
x <- array(c(18, 12, 12, 8, 2, 8, 8, 32), dim = c(2, 2, 2),
           dimnames = list("Treatment" = c("A", "B"),
                           "Response" = c("Success", "Failure"),
                           "Clinic" = c("1", "2")))
x
apply(x, 3, twoby2)[[c(1, 2)]]
twoby2(x[,,1] + x[,,2])
```
We can see here that treatment and response variables are conditionally independent
(have odds ratio of 1, controlling for clinic), but the marginal odds ratio is 2.
The reason for this stems from the fact that the conditional odds of receiving
treatment A are six times higher at clinic 1 than 2.